---
title: "Exercise 7 lkm"
author: "Lola Kourouma"
date: "11/3/2020"
output: html_document
---

## Exercise 7: playing with design/contrast matrices, UpSet plots and differential expression with count data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library("limma")
library("affy")
library("preprocessCore")
library("UpSetR")
```

### Question 1. Using the estrogen dataset from Exercise 5, create an UpSet plot summarizing the numbers of differentially expressed genes for the 3 contrasts: ‘E10’, ‘E48’ and ‘Time’ (contrasts as defined in Exercise 5 using makeContrasts()). Hint: you can use decideTests() to summarize the results, given some thresholds. UpSet plots are the new Venn diagrams and can be created with the UpSetR package.

```{r}
unzip("affy_estrogen.zip")
ddir <- "affy_estrogen"
dir(ddir) 
```

```{r}
# preprocess affymetrix data
targets <- readTargets("targets.txt", path=ddir)
targets
```

```{r, message=FALSE}
abatch <- ReadAffy(filenames=targets$filename,
                   celfile.path=ddir)
eset <- rma(abatch)  # bg correct, normalize, summarize

```

```{r, message=FALSE}
# do the limma modeling
f <- paste(targets$estrogen,targets$time.h,sep="")
f <- factor(f)

# create design matrix
design <- model.matrix(~0+f)
colnames(design) <- levels(f)
design
```

```{r}
fit <- lmFit(eset, design)

cont.matrix <- makeContrasts(E10="present10-absent10",
                             E48="present48-absent48",
                             Time="absent48-absent10",levels=design)
cont.matrix
```

```{r}
fit2  <- contrasts.fit(fit, cont.matrix)
fit2  <- eBayes(fit2)      #compute t-stat, F-stat, log-odds
class(fit2)
```

```{r}
#summarizing the results using decideTests()

summ <- decideTests(object = fit2, p.value = 0.05)

summ <- (abs(summ))
            
head(summ, 10)

```

```{r}
#generating an UpSet plot

summdata <- as.data.frame(summ)
str(summdata)

upset(summdata)

```

### Question 3. Fix the code above to work on the count table that was already read in. Add in some spot checks, including an MDS plot from plotMDS() (or similar), a look at the dispersion-mean plot using plotBCV() (or similar) and a look at the overall M vs A plot using plotSmear() (or similar).

```{r, message=FALSE}

library("edgeR")
samples <- read.table("samples.txt", header=TRUE,
                      row.names=5, stringsAsFactors=FALSE)
samples

```

```{r}
counts <- readDGE(samples$countfile)$counts
head(counts)
```

```{r}
(grp <- gsub("\\-.[0-9]*","",colnames(counts)))

```

```{r}

group <- factor(c(1,1,1,2,2,2,2))

y <- DGEList(counts=counts,group=group)

y <- calcNormFactors(y)

y$samples
```

```{r}
#design matrix 
des <- model.matrix(~group)
des
```

```{r}

y <- estimateDisp(y,des)
#To perform likelihood ratio tests:
fit_ <- glmFit(y,des)
lrt <- glmLRT(fit_,coef=2)
topTags(lrt)

```

#### MDS plot:
```{r}
num.group <- as.numeric(group)
plotMDS(y, top=200, labels = samples$rep, col=num.group)

```


#### Dispersion-Mean-Plot:
```{r}

plotBCV(y)

```


#### Overall M vs A plot 
```{r}

plotSmear(y)

```


### Question 4. For the MDS plot, investigate what happens differently when you call plotMDS() on the DGEList object and when you call it on the normalized counts – for normalized counts, see the cpm() function. Explain why it is different.

```{r}

norm_counts <- cpm(counts)
head(norm_counts)

plotMDS(norm_counts, top=200, labels = samples$rep, col=num.group)

```


### Question 5. Again, the above code for voom is not run here. Fix this code to work with the count dataset above and then do some ``spot checks’’ of the genes called DE. For example, make a plot of the plot of estimated fold changes for the two methods (limma and edgeR), compare the P-values of the two methods, or create a venn diagram of the called sets of DE genes at a set threshold.

```{r}

v <- voom(y, design=des, plot = TRUE)
vf = lmFit(v, design = des) 
vf = eBayes(vf)
topTable(vf, coef=2)

```

#### Spot check of the genes called DE:

```{r}
pv <- vf$p.value
str(pv)
```

```{r}
plrt <- lrt$table$PValue
str(plrt)
```

```{r}

plot(plrt, pv[,2], xlab = "P edgeR", ylab = "P Voom")

```


